---
title: "ASAP invitro systems differential expression analysis of TEs"
author: "Raquel Garza"
output:
  html_document:
    df_print: paged
---

Here we perform differential expression analysis (DEA) of ERVs for our in vitro experiments using microglia or dopaminergic neurons and different treatments (see metadata).

Before we start, I load libraries and define some helper functions to visualize the effect as a mean plot (credits to Per Brattås! who wrote the original code for it)

```{r class.source = 'fold-hide'}
## getSignName ##
# Get significantly different gene names. 
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# x = results object from deseq
# p = padj threshold for significance
# l = log2FC threshold for significance
getSignName <- function(x,p,l=0) {
  up <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange > l,]
  down <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange < -l,]
  return(list(up=rownames(up),down=rownames(down)))
}
## getAverage ##
# Get average expression (normalized by median of ratios) of each of the conditions in a deseq object.
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# dds = deseq object
getAverage <- function(dds) {
  baseMeanPerLvl <- sapply( levels(dds$condition), function(lvl) rowMeans( counts(dds,normalized=TRUE)[,dds$condition == lvl] ) )
  baseSDPerLvl <- sapply( levels(dds$condition), function(lvl) apply( counts(dds,normalized=TRUE)[,dds$condition == lvl],1,sd ) )
  colnames(baseSDPerLvl) <- paste("st.dev:",colnames(baseSDPerLvl),sep="")
  return(list(Mean=baseMeanPerLvl,SD=baseSDPerLvl))
}

meanPlot_cus <- function(exp,test,c1 = "condition 1",c2 = "condition 2",p=.05,l=0,id=F, ttl="", 
                         repel=TRUE, col1="tomato", col2="darkturquoise", col3="grey", highlights=NA){
  sign <- getSignName(x = test,p = p,l = l)
  u <- sign$up
  d <- sign$down
  
  #color up and down sign..
  colVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = col1,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = col2, no =col3))
  colVec[is.na(colVec)] <- "steelblue" ## if NA make sure it's not counted as <p
  #size of points
  cexVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = 0.35,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = 0.35, no = 0.3))
  
  exp_log <- as.data.frame(log2(exp[,c(c1, c2)] + 0.5))
  exp_log$Name <- rownames(exp_log)
  
  exp_log$reg <- factor(ifelse(exp_log$Name %in% u, paste('upregulated in ', c1, ' (', length(u), ')', sep =''),
                               ifelse(exp_log$Name %in% d, paste('downregulated in ', c1,' (', length(d), ')', sep =''), paste('not significant', ' (', (nrow(test) - length(u) - length(d)), ')', sep=''))))
  
  library(ggrepel)
  if(repel == TRUE){
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), label=Name, color=reg)) + geom_point(aes(size=cexVec), alpha=0.7)+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")+ geom_text_repel(data = subset(exp_log, Name %in% u | Name %in% d),direction    = "y", nudge_y = 0.4, nudge_x = -0.5)
  }
  else{
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), color=reg)) + geom_point(aes(size=cexVec), alpha=0.7)+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")
  }
  plt <- plt + labs(x=paste("log2(mean ",c2,")",sep=""), 
                    y=paste("log2(mean ",c1,")",sep=""),
                    title=paste(ttl, paste(c1," vs. ",c2,sep=""), sep = ': '),
                    subtitle=paste("p-adj < ",p,", log2(fc) > ",l,sep=""))+ theme_bw() +
    theme( plot.title = element_text( size=14, face="bold"), text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank(), legend.title=element_blank()) 
  
  
  if(id==T) {
    
    identify(log2(exp[,1]),log2(exp[,2]),labels = rownames(exp))
    
  }
  
  if(!is.na(highlights)){
    plt <- plt + geom_point(data=exp_log[highlights,], aes(x=get(c2), y=get(c1)), colour="springgreen3", size=5, shape=1, stroke=2)
  }
  return(plt)
  
}
```

## Loading ERV (retrotector) count matrices

```{r}
library(data.table)
library(ggplot2)
library(ggpubr)
library(stringr)
library(DESeq2)
library(bedr)
library(openxlsx)
library(pheatmap)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(openxlsx)
library(ggVennDiagram)
library(UpSetR)

samplesheet <- fread("/Volumes/MyPassport/ASAP/code/ASAP_invitro/original/samplesheet.tab", data.table = F, header = T)
samplesheet_list <- split(samplesheet, f = list(samplesheet$celltype, samplesheet$cellline, samplesheet$condition))
samplesheet_list <- samplesheet_list[sapply(samplesheet_list, nrow) > 0]

# Split samplesheet per cell lines and treatments
celltypes <- unique(sapply(str_split(names(samplesheet_list), "[[.]]"), `[[`, 1))
for(celltype in celltypes){
  treatments_celltype <- names(samplesheet_list)[grepl(celltype, names(samplesheet_list))]
  celllines <- unique(sapply(str_split(treatments_celltype, "[[.]]"), `[[`, 2))
  for(cellline in celllines){
    celllines_treatments_celltype <- names(samplesheet_list)[grepl(celltype, names(samplesheet_list)) & grepl(cellline, names(samplesheet_list))]
    celllines_treatments_celltype <- celllines_treatments_celltype[!grepl("ctrl", celllines_treatments_celltype)]
    for(treatment in celllines_treatments_celltype){
      corresponding_ctrl <- paste(celltype, cellline, "ctrl", sep=".")
      samplesheet_list[[treatment]] = rbind(samplesheet_list[[treatment]], samplesheet_list[[corresponding_ctrl]])
    }
    samplesheet_list[[corresponding_ctrl]] = NULL
  }
}

# Read gene count matrices
# write.xlsx(samplesheet_list, "/Volumes/MyPassport/ASAP_invitro/results/tables/samplesheet_comparisons.xlsx")
path <- "/Volumes/MyPassport/ASAP/code/ASAP_invitro/original/TEcounts/unique/"
# samples <- samplesheet$samples
samples <- unique(do.call(rbind, samplesheet_list))$sample
# Rerun featurecounts to the organoids 
for(i in 1:length(samples)){
  sample <- samples[i]
  if(i == 1){
    herv_counts <- fread(paste(path, sample, "_ERV_count_matrix_2.csv", sep=""), data.table = F)    
    colnames(herv_counts)[ncol(herv_counts)] <- sample
    rownames(herv_counts) <- herv_counts$hervid
    row_order <- rownames(herv_counts)
  }else{
    tmp <- fread(paste(path, sample, "_ERV_count_matrix_2.csv", sep=""), data.table = F)
    colnames(tmp)[ncol(tmp)] <- sample
    rownames(tmp) <- tmp$hervid
    herv_counts <- cbind(herv_counts[row_order,], tmp[row_order,sample,drop=F])
  }
}
```

## Remove outlier samples

```{r}
samplesheet <- samplesheet[which(samplesheet$sample != "ASAP_PolyIC_Rep2_bulkRNA_vMBneurons_RC17_day52_S26"),]
samplesheet_list$neurons.RC17.PolyIC <- samplesheet_list$neurons.RC17.PolyIC[which(samplesheet_list$neurons.RC17.PolyIC$sample != "ASAP_PolyIC_Rep2_bulkRNA_vMBneurons_RC17_day52_S26"),]

rownames(samplesheet) <- samplesheet$sample
rownames(herv_counts) <- (herv_counts$Geneid)
rownames(samplesheet) <- samplesheet$sample
```

## ERV Differential Expression Analysis (DEA)

Per experiment
```{r}
herv_dds_list <- list()
herv_res_list <- list()
herv_res_list_df <- list()
herv_exp_list <- list()
herv_vst_list <- list()

for(i in names(samplesheet_list)){
  rownames(samplesheet_list[[i]]) <- samplesheet_list[[i]]$sample
  print(i)
  conditions <- unique(samplesheet_list[[i]]$condition)
  print(c("Conditions: ", conditions))

  effect <- conditions[which(conditions != "ctrl")]
  test <- paste("condition_", effect ,"_vs_ctrl", sep = "")
  print(c("Test: ", test))
  counts <- herv_counts[,rownames(samplesheet_list[[i]])]
  herv_dds_list[[i]] <- DESeqDataSetFromMatrix(counts, samplesheet_list[[i]], design =  ~ condition)
  herv_dds_list[[i]]$condition <- relevel(herv_dds_list[[i]]$condition, "ctrl")
  herv_dds_list[[i]] <- DESeq(herv_dds_list[[i]])
  herv_res_list[[i]] <- lfcShrink(herv_dds_list[[i]], test)
  herv_exp_list[[i]] <- getAverage(herv_dds_list[[i]])
  herv_res_list_df[[i]] <- as.data.frame(herv_res_list[[i]])
  herv_res_list_df[[i]]$herv_name <- rownames(herv_res_list_df[[i]])
  herv_vst_list[[i]] <- varianceStabilizingTransformation(herv_dds_list[[i]])
}

```

## PCA plots

```{r}
plotPCA(herv_vst_list$microglia.H9.IFNg, intgroup="condition") + ggtitle("Microglia H9 IFNg: Condition") + theme_bw()
plotPCA(herv_vst_list$microglia.H9.TNFa, intgroup="condition") + ggtitle("Microglia H9 TNFa: Condition") + theme_bw()
plotPCA(herv_vst_list$microglia.H9.PolyIC, intgroup="condition") + ggtitle("Microglia H9 PolyIC: Condition") + theme_bw()
plotPCA(herv_vst_list$neurons.KOLF.IFNg, intgroup="condition") + ggtitle("Neurons KOLF IFNg: Condition") + theme_bw()
plotPCA(herv_vst_list$neurons.RC17.IFNg, intgroup="condition") + ggtitle("Neurons RC17 IFNg: Condition") + theme_bw()
plotPCA(herv_vst_list$neurons.KOLF.TNFa, intgroup="condition") + ggtitle("Neurons KOLF TNFa: Condition") + theme_bw()
plotPCA(herv_vst_list$neurons.RC17.TNFa, intgroup="condition") + ggtitle("Neurons RC17 TNFa: Condition") + theme_bw()
plotPCA(herv_vst_list$neurons.KOLF.PolyIC, intgroup="condition") + ggtitle("Neurons KOLF PolyIC: Condition") + theme_bw()
plotPCA(herv_vst_list$neurons.RC17.PolyIC, intgroup="condition") + ggtitle("Neurons RC17 PolyIC: Condition") + theme_bw()
```

## Visualization

Mean plots, volcano plots and excel tables...
```{r}
library(EnhancedVolcano)
# openxlsx::write.xlsx(herv_res_list_df, file = "/Volumes/MyPassport/ASAP_invitro/results/tables/hervs_DEA.xlsx")

volcano_plots_list <- list()
mean_plots_list <- list()

for(i in names(herv_res_list)){
  tmp <- EnhancedVolcano(herv_res_list[[i]], 
                                             drawConnectors = T, 
                                             lab = NA, 
                                             x = 'log2FoldChange',
                                             y = 'padj',
                                             pCutoff = 0.05,
                                             FCcutoff = 1)
  tmp$data$Sig <- ifelse(tmp$data$log2FoldChange > 1 & tmp$data$Sig == "FC_P", "Upregulated", 
                                             ifelse(tmp$data$log2FoldChange < -1 & tmp$data$Sig == "FC_P", "Downregulated", "Not significant"))
  tmp$data$Sig <- paste(tmp$data$Sig, table(tmp$data$Sig)[tmp$data$Sig], sep = ": ")
  
  not_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Not significant")]
  up_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Upregulated")]
  down_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Downregulated")]
  
  point_colours <- c("grey","darkturquoise", "tomato")
  names(point_colours) <- c(not_sig_lab, down_sig_lab, up_sig_lab)
  point_sizes <- c(1,2,2)
  names(point_sizes) <- c(not_sig_lab, down_sig_lab, up_sig_lab)  
  
  title <- str_replace_all(i, pattern = "_", replacement = " ")
  
  tmp <- tmp$data
  volcano_plots_list[[i]] <- ggplot(tmp, aes(x=log2FoldChange, y=-log10(padj), colour=Sig, size=Sig)) + geom_point(alpha=0.7) + 
    geom_vline(xintercept = 1, linetype="dotted") + geom_vline(xintercept = -1, linetype="dotted") + 
    geom_hline(yintercept = -log10(0.05), linetype="dotted") + labs(colour="") + 
    scale_color_manual(values = point_colours)  + 
    scale_size_manual(values = point_sizes, guide="none") +
    theme_bw() +
    theme(text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
    ggtitle(title)
  
  effect <- unique(samplesheet_list[[i]][which(samplesheet_list[[i]]$condition != "ctrl"),"condition"])
  mean_plots_list[[i]] <- meanPlot_cus(exp = herv_exp_list[[i]]$Mean, 
                                       test = herv_res_list[[i]], #col1 = "tomato", col2 = "grey", col3="blue",
                                       c1 = effect, c2="ctrl", p = 0.05, l=1, ttl = title, repel = F)
    
}


volcano_plots_list
mean_plots_list

```

## How well does the effect correlate between the two cell lines? (Neurons)

```{r}
herv_res_neurons <- data.frame("herv" = rownames(herv_res_list_df$microglia.H9.IFNg),
                               "log2FC_neurons.RC17.IFNg" = herv_res_list_df$neurons.RC17.IFNg[rownames(herv_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.KOLF.IFNg" = herv_res_list_df$neurons.KOLF.IFNg[rownames(herv_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.RC17.PolyIC" = herv_res_list_df$neurons.RC17.PolyIC[rownames(herv_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.KOLF.PolyIC" = herv_res_list_df$neurons.KOLF.PolyIC[rownames(herv_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.RC17.TNFa" = herv_res_list_df$neurons.RC17.TNFa[rownames(herv_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.KOLF.TNFa" = herv_res_list_df$neurons.KOLF.TNFa[rownames(herv_res_list_df$microglia.H9.IFNg), "log2FoldChange"])

ggarrange(
  herv_res_neurons %>% 
  drop_na(log2FC_neurons.RC17.IFNg, log2FC_neurons.KOLF.IFNg) %>% 
ggplot(aes(x=log2FC_neurons.RC17.IFNg, y=log2FC_neurons.KOLF.IFNg)) + 
  geom_point(alpha=0.5) + theme_bw() + 
  geom_hline(yintercept = 0, linetype="dashed", color="lightgrey") + 
  geom_vline(xintercept = 0, linetype="dashed", color="lightgrey") + 
  labs(y = "log2FC KOLF IFNg", x = "log2FC RC17 IFNg") + geom_smooth(method = "lm", color="darkolivegreen2") + 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) + ggtitle("Neurons + IFNg"),

herv_res_neurons %>% 
  drop_na(log2FC_neurons.RC17.TNFa, log2FC_neurons.KOLF.TNFa) %>% 
ggplot(aes(x=log2FC_neurons.RC17.TNFa, y=log2FC_neurons.KOLF.TNFa)) + 
  geom_point(alpha=0.5) + theme_bw() + 
  geom_hline(yintercept = 0, linetype="dashed", color="lightgrey") + 
  geom_vline(xintercept = 0, linetype="dashed", color="lightgrey") + 
  labs(y = "log2FC KOLF TNFa", x = "log2FC RC17 TNFa") + geom_smooth(method = "lm", color="darkolivegreen2") + 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) + ggtitle("Neurons + TNFa"),

herv_res_neurons %>% 
  drop_na(log2FC_neurons.RC17.PolyIC, log2FC_neurons.KOLF.PolyIC) %>% 
ggplot(aes(x=log2FC_neurons.RC17.PolyIC, y=log2FC_neurons.KOLF.PolyIC)) + 
  geom_point(alpha=0.5) + theme_bw() + 
  geom_hline(yintercept = 0, linetype="dashed", color="lightgrey") + 
  geom_vline(xintercept = 0, linetype="dashed", color="lightgrey") + 
  labs(y = "log2FC KOLF PolyIC", x = "log2FC RC17 PolyIC") + geom_smooth(method = "lm", color="darkolivegreen2") + 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) + ggtitle("Neurons + PolyIC"), ncol=3)

```

## What is the overlap of upregulated ERVs between the three treatments?

```{r}
venn_neurons <- list("neurons.KOLF.IFNg" = getSignName(herv_res_list_df$neurons.KOLF.IFNg, p = 0.05, l = 1)$up,
                    "neurons.RC17.IFNg" = getSignName(herv_res_list_df$neurons.RC17.IFNg, p = 0.05, l = 1)$up,
                    "neurons.KOLF.TNFa" = getSignName(herv_res_list_df$neurons.KOLF.TNFa, p = 0.05, l = 1)$up)

# write.xlsx(venn_neurons, "/Volumes/MyPassport/ASAP_invitro/results/tables/upreg_HERV_neurons.xlsx")
upset(fromList(venn_neurons), nsets = 100, nintersects = 100)

venn_microglia <- list("microglia.H9.IFNg" = getSignName(herv_res_list_df$microglia.H9.IFNg, p = 0.05, l = 1)$up,
                    "microglia.H9.TNFa" = getSignName(herv_res_list_df$microglia.H9.TNFa, p = 0.05, l = 1)$up,
                    "microglia.H9.PolyIC" = getSignName(herv_res_list_df$microglia.H9.PolyIC, p = 0.05, l = 1)$up)

# write.xlsx(venn_microglia, "/Volumes/MyPassport/ASAP_invitro/results/tables/upset_HERV_upreg_microglia_IFNg_TNFa_PolyIC.xlsx")

upset(fromList(venn_microglia), nsets = 100, nintersects = 100)
```


## Visualization of upregulated ERVs on heatmap

Normalization using gene-sizeFactors

```{r}
sizefactors <- read.table("/Volumes/MyPassport/ASAP/code/ASAP_invitro/original/results/tables/gene_sizeFactors.tab")
colnames(sizefactors) <- c("sample", "sizeFactor")
rownames(sizefactors) <- sizefactors$sample

herv_counts_norm <- herv_counts[,samples]
herv_counts_norm[] <- mapply('/', herv_counts_norm[,samples], sizefactors[samples,"sizeFactor"])

herv_counts_norm <- herv_counts_norm[which(rowSums(herv_counts_norm) > 0),]
herv_counts_norm$TE_id <- rownames(herv_counts_norm)
herv_counts_norm_melt <- reshape2::melt(herv_counts_norm)

herv_counts_norm_melt <- merge(herv_counts_norm_melt, samplesheet[,c("sample", "condition", "celltype")], by.x="variable", by.y="sample")
microglia_samples <- c(samplesheet_list$microglia.H9.IFNg[, "sample"],
                       samplesheet_list$microglia.H9.TNFa[, "sample"],
                       samplesheet_list$microglia.H9.PolyIC[, "sample"])
microglia_samples <- microglia_samples[order(samplesheet[microglia_samples,"condition"])]

list_to_tidy_df <- function(x){
  df <- as.data.frame(venn_microglia[[x]])
  colnames(df) <- "TE_id"
  df$type <- x
  return(df)
}

venn_microglia_tidy_df <- do.call(rbind, lapply(names(venn_microglia), FUN = list_to_tidy_df))
venn_microglia_tidy_df <- venn_microglia_tidy_df %>%   
  group_by(TE_id) %>%
  summarise(across(everything(), ~paste0(unique(.), collapse = "|"))) %>% 
  as.data.frame()
rownames(venn_microglia_tidy_df) <- venn_microglia_tidy_df$TE_id

# This file contains the subfamily annotation
hg38_retro_names <- fread("/Volumes/MyPassport/annotations/retrotector/hg38_db_mysql/hg38rt102v150418.bed", data.table=F, skip=1)
colnames(hg38_retro_names) <- c("chr", "start", "end", "ID", "score", "strand", "TE_info")
# We will use the coordinates as ID to merge with the annotation file we have been using (to get the HERV_dupXXXX)
hg38_retro_names$coords <- paste(hg38_retro_names$chr, paste(hg38_retro_names$start, hg38_retro_names$end, sep="-"), sep=":")
hg38_retro_names <- hg38_retro_names %>%   
  group_by(coords) %>%
  summarise(across(everything(), ~paste0(unique(.), collapse = "|"))) %>% # "Remove" (collapse) duplicates
  as.data.frame()

hg38_retro <- fread("/Volumes/MyPassport/annotations/human/retrotector/hg38_HERVs_prediction.bed", data.table=F, skip=1)
colnames(hg38_retro) <- c("chr", "start", "end", "dot", "strand", "dot2", "TE_id")
hg38_retro$coords <- paste(hg38_retro$chr, paste(hg38_retro$start, hg38_retro$end, sep="-"), sep=":")

# table(unique(hg38_retro_names$coords) %in% unique(hg38_retro$coords)) # There are some HERVs in alternative scaffolds here, so these would go on FALSE
# But it doesnt matter for the 34 elements we need to plot... 
hg38_retro <- merge(hg38_retro, hg38_retro_names[,c("coords", "TE_info", "score")], by="coords") # merge to get HERV_dupXXXX IDs
rownames(hg38_retro) <- hg38_retro$TE_id
annotate_rows <- hg38_retro[which(hg38_retro$TE_id %in% venn_microglia_tidy_df$TE_id),c("TE_info", "coords"), drop=F]
annotate_rows <- cbind(annotate_rows[rownames(annotate_rows),], venn_microglia_tidy_df[rownames(annotate_rows),"type", drop = F])
colnames(annotate_rows) <- c("TE_info", "coords", "type")

annotate_rows$TE_subfam <- ifelse(grepl("retrovirus W", annotate_rows$TE_info), "HERVW",
       ifelse(grepl("HERV-K", annotate_rows$TE_info), "HERVK",
              ifelse(grepl("RTVL-H2", annotate_rows$TE_info), "RTVL-H2",
                     ifelse(grepl("ERV9", annotate_rows$TE_info), "ERV9",
                            ifelse(grepl("ERV3", annotate_rows$TE_info), "ERV3", annotate_rows$coords)))))


pheatmap(herv_counts_norm[venn_microglia_tidy_df$TE_id,microglia_samples], 
         show_colnames = F, show_rownames = T, scale = "row", cluster_cols = F,
         annotation_row = annotate_rows[,c("TE_subfam", "type")], annotation_colors = list("condition" = c("ctrl" = "lightgrey", 
                                                                                                           "IFNg" = "#31c4b5",
                                                                                                           "PolyIC" = "#ea7792",
                                                                                                           "TNFa" = "#ad9dcc")),
         annotation_col = samplesheet[microglia_samples,"condition",drop=F], main = "Microglia IFNg, PolyIC, TNFa")

```

