---
title: ASAP invitro cell cultures + treatments. Differential expression analysis of
  genes.
author: "Raquel Garza"
output:
  html_document:
    df_print: paged
---

Here we perform differential expression analysis (DEA) of genes for our in vitro experiments using microglia, astrocytes or dopaminergic neurons and different treatments (see metadata).

Before we start, I load libraries and define some helper functions to visualize the effect as a mean plot (credits to Per Brattås! who wrote the original code for it)

```{r class.source = 'fold-hide'}
## getSignName ##
# Get significantly different gene names. 
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# x = results object from deseq
# p = padj threshold for significance
# l = log2FC threshold for significance
getSignName <- function(x,p,l=0) {
  up <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange > l,]
  down <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange < -l,]
  return(list(up=rownames(up),down=rownames(down)))
}
## getAverage ##
# Get average expression (normalized by median of ratios) of each of the conditions in a deseq object.
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# dds = deseq object
getAverage <- function(dds) {
  baseMeanPerLvl <- sapply( levels(dds$condition), function(lvl) rowMeans( counts(dds,normalized=TRUE)[,dds$condition == lvl] ) )
  baseSDPerLvl <- sapply( levels(dds$condition), function(lvl) apply( counts(dds,normalized=TRUE)[,dds$condition == lvl],1,sd ) )
  colnames(baseSDPerLvl) <- paste("st.dev:",colnames(baseSDPerLvl),sep="")
  return(list(Mean=baseMeanPerLvl,SD=baseSDPerLvl))
}

meanPlot_cus <- function(exp,test,c1 = "condition 1",c2 = "condition 2",p=.05,l=0,id=F, ttl="", 
                         repel=TRUE, col1="tomato", col2="darkturquoise", col3="grey", highlights=NA){
  sign <- getSignName(x = test,p = p,l = l)
  u <- sign$up
  d <- sign$down
  
  #color up and down sign..
  colVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = col1,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = col2, no =col3))
  colVec[is.na(colVec)] <- "steelblue" ## if NA make sure it's not counted as <p
  #size of points
  cexVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = 0.35,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = 0.35, no = 0.3))
  
  exp_log <- as.data.frame(log2(exp[,c(c1, c2)] + 0.5))
  exp_log$Name <- rownames(exp_log)
  
  exp_log$reg <- factor(ifelse(exp_log$Name %in% u, paste('upregulated in ', c1, ' (', length(u), ')', sep =''),
                               ifelse(exp_log$Name %in% d, paste('downregulated in ', c1,' (', length(d), ')', sep =''), paste('not significant', ' (', (nrow(test) - length(u) - length(d)), ')', sep=''))))
  
  library(ggrepel)
  if(repel == TRUE){
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), label=Name, color=reg)) + geom_point(aes(size=cexVec), alpha=0.7)+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")+ geom_text_repel(data = subset(exp_log, Name %in% u | Name %in% d),direction    = "y", nudge_y = 0.4, nudge_x = -0.5)
  }else{
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), color=reg)) + geom_point(aes(size=cexVec), alpha=0.7)+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")
  }
  plt <- plt + labs(x=paste("log2(mean ",c2,")",sep=""), 
                    y=paste("log2(mean ",c1,")",sep=""),
                    title=paste(ttl, paste(c1," vs. ",c2,sep=""), sep = ': '),
                    subtitle=paste("p-adj < ",p,", log2(fc) > ",l,sep=""))+ theme_bw() +
    theme( plot.title = element_text( size=14, face="bold"), text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank(), legend.title=element_blank()) 
  
  
  if(id==T) {
    identify(log2(exp[,1]),log2(exp[,2]),labels = rownames(exp))
  }
  
  if(!all(is.na(highlights))){
    plt <- plt + geom_text_repel(data = exp_log[highlights,], aes(label=Name), size = 3, segment.size= 0.1, max.overlaps = 100, color = "black") #+ 
      #geom_point(data=exp_log[highlights,], aes(x=get(c2), y=get(c1)), colour="springgreen3", size=5, shape=1, stroke=2)
  }
  return(plt)
  
}
```

## Loading gene count matrices
```{r}
library(data.table)
library(ggplot2)
library(ggpubr)
library(stringr)
library(DESeq2)
library(openxlsx)
library(tidyverse)
library(EnhancedVolcano)
library(bedr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pheatmap)
library(dplyr)
library(UpSetR)

# Gene name annotation
gene_annotation <- fread("/Volumes/MyPassport/annotations/human/gencode/v38/gencode.v38.annotation.gene_names.tab", data.table = F, skip=1, header = F)
colnames(gene_annotation) <- c("Geneid", "gene_name", "gene_type")

# Samplesheet / metadata
samplesheet <- fread("/Volumes/MyPassport/ASAP/code/ASAP_invitro/original/samplesheet.tab", data.table = F, header = T)
samplesheet_list <- split(samplesheet, f = list(samplesheet$celltype, samplesheet$cellline, samplesheet$condition))
samplesheet_list <- samplesheet_list[sapply(samplesheet_list, nrow) > 0]

# Split samplesheet per cell lines and treatments
celltypes <- unique(sapply(str_split(names(samplesheet_list), "[[.]]"), `[[`, 1))
for(celltype in celltypes){
  treatments_celltype <- names(samplesheet_list)[grepl(celltype, names(samplesheet_list))]
  celllines <- unique(sapply(str_split(treatments_celltype, "[[.]]"), `[[`, 2))
  for(cellline in celllines){
    celllines_treatments_celltype <- names(samplesheet_list)[grepl(celltype, names(samplesheet_list)) & grepl(cellline, names(samplesheet_list))]
    celllines_treatments_celltype <- celllines_treatments_celltype[!grepl("ctrl", celllines_treatments_celltype)]
    for(treatment in celllines_treatments_celltype){
      corresponding_ctrl <- paste(celltype, cellline, "ctrl", sep=".")
      samplesheet_list[[treatment]] = rbind(samplesheet_list[[treatment]], samplesheet_list[[corresponding_ctrl]])
    }
    samplesheet_list[[corresponding_ctrl]] = NULL
  }
}

# Read gene count matrices
# write.xlsx(samplesheet_list, "/Volumes/MyPassport/ASAP_invitro/results/tables/samplesheet_comparisons.xlsx")
path <- "/Volumes/MyPassport/ASAP/code/ASAP_invitro/original/gene_counts/unique/"
# samples <- samplesheet$samples
samples <- unique(do.call(rbind, samplesheet_list))$sample
for(i in 1:length(samples)){
  sample <- samples[i]
  if(i == 1){
    gene_counts <- fread(paste(path, sample, "_gene_count_matrix_2.csv", sep=""), data.table = F)    
    colnames(gene_counts)[ncol(gene_counts)] <- sample
    rownames(gene_counts) <- gene_counts$Geneid
    row_order <- rownames(gene_counts)
  }else{
    tmp <- fread(paste(path, sample, "_gene_count_matrix_2.csv", sep=""), data.table = F)
    colnames(tmp)[ncol(tmp)] <- sample
    rownames(tmp) <- tmp$Geneid
    gene_counts <- cbind(gene_counts[row_order,], tmp[row_order,sample,drop=F])
  }
}
```

## Cell type markers

Normalization using sizeFactors calculated with all samples. 
```{r}
samplesheet <- samplesheet[which(samplesheet$sample != "ASAP_PolyIC_Rep2_bulkRNA_vMBneurons_RC17_day52_S26"),] # Outlier in PCA and markers
samplesheet_list$neurons.RC17.PolyIC <- samplesheet_list$neurons.RC17.PolyIC[which(samplesheet_list$neurons.RC17.PolyIC$sample != "ASAP_PolyIC_Rep2_bulkRNA_vMBneurons_RC17_day52_S26"),] # Outlier in PCA and markers

rownames(samplesheet) <- samplesheet$sample
rownames(gene_counts) <- (gene_counts$Geneid)
# Differential gene expression analysis
gene_dds <- DESeqDataSetFromMatrix(gene_counts[,rownames(samplesheet)], samplesheet, design =  ~ condition)
gene_dds$condition <- relevel(gene_dds$condition, "ctrl")
gene_dds <- DESeq(gene_dds)
# Size factors for TE normalization
# write.table(as.data.frame(gene_dds$sizeFactor), file = "/Volumes/MyPassport/ASAP_invitro/results/tables/gene_sizeFactors.tab", quote = F, sep="\t", col.names = F)

# Normalized counts to plot markers
gene_count_norm <- counts(gene_dds, normalized=T)
neuronal_genes <- c("WNT3", "POU3F2", "NEUROD4", "NEUROG2", "FOXG1", "FOXA2", "LMX1A", "NEUROD1", "SOX11", "NES", "SOX1", "SOX2", "ASCL1", "DCX", "NCAM1", "MAP2", "PPP1R1B", "DRD1", "DRD2")
samplesheet <- samplesheet[order(samplesheet$cellline, samplesheet$celltype),]
samplesheet_rc17_neurons_ctrl <- samplesheet[which(samplesheet$celltype == "neurons" & samplesheet$condition == "ctrl" & samplesheet$cellline == "RC17"),]

pheatmap(log2(gene_count_norm[neuronal_genes,rownames(samplesheet_rc17_neurons_ctrl)]+0.5), cluster_cols = F, cluster_rows = F, scale = "none", main = "All samples", annotation_col = samplesheet[,c("cellline", "celltype", "condition")])

pheatmap(log2(gene_count_norm[neuronal_genes,rownames(samplesheet)]+0.5), cluster_cols = F, cluster_rows = F, scale = "none", main = "All samples", annotation_col = samplesheet[,c("cellline", "celltype", "condition")])

pheatmap(log2(gene_count_norm[neuronal_genes,rownames(samplesheet)]+0.5), cluster_cols = F, cluster_rows = F, scale = "row", main = "All samples", annotation_col = samplesheet[,c("cellline", "celltype", "condition")])

# Subset just the neuron samples
neuron_samples <- c(samplesheet_list$neurons.RC17.IFNg$sample,
                    samplesheet_list$neurons.KOLF.IFNg$sample,
                    samplesheet_list$neurons.RC17.TNFa$sample,
                    samplesheet_list$neurons.KOLF.TNFa$sample,
                    samplesheet_list$neurons.RC17.PolyIC$sample,
                    samplesheet_list$neurons.KOLF.PolyIC$sample)
rownames(samplesheet) <- samplesheet$sample

pheatmap(log2(gene_count_norm[c(neuronal_genes, "RBFOX3", "PGP", "GRIN1", "GAD1", "TH", "HS3ST2", "CALB2", "CNR1"),neuron_samples]+0.5), 
         cluster_cols = F, cluster_rows = F, scale = "none", 
         main = "Neurons", annotation_col = samplesheet[neuron_samples,c("cellline", "celltype", "condition")])

# Subset just the microglia samples
microglia_samples <- c(samplesheet_list$microglia.H9.IFNg$sample,
                    samplesheet_list$microglia.H9.TNFa$sample,
                    samplesheet_list$microglia.H9.PolyIC$sample)
microglia_samples <- microglia_samples[which(grepl("non_treated", microglia_samples))]

microglial_genes <- c("C1QA", "CD200R1", "CSF1R", "CX3CR1", "GAS6", "GPR34", "IL1B", "CXCL8", "MERTK", "OLFML3", "P2RY12", "PROS1", "TGFB1", "TMEM119", "TNF", "TREM2")
pheatmap(log2(gene_count_norm[c(microglial_genes),microglia_samples]+0.5), 
         cluster_cols = F, cluster_rows = T, scale = "none", 
         main = "Microglia", annotation_col = samplesheet[microglia_samples,c("celltype", "condition")])
```

## Gene Differential Expression Analysis (DEA)

Looped per experiment
```{r}
gene_dds_list <- list()
gene_res_list <- list()
gene_res_list_df <- list()
gene_exp_list <- list()
gene_vst_list <- list()
rownames(gene_counts) <- gene_counts$Geneid

for(i in names(samplesheet_list)){
  rownames(samplesheet_list[[i]]) <- samplesheet_list[[i]]$sample
  print(i)
  conditions <- unique(samplesheet_list[[i]]$condition)
  print(c("Conditions: ", conditions))

  effect <- conditions[which(conditions != "ctrl")]
  test <- paste("condition_", effect ,"_vs_ctrl", sep = "")
  print(c("Test: ", test))
  counts <- gene_counts[,rownames(samplesheet_list[[i]])]
  gene_dds_list[[i]] <- DESeqDataSetFromMatrix(counts, samplesheet_list[[i]], design =  ~ condition)
  gene_dds_list[[i]]$condition <- relevel(gene_dds_list[[i]]$condition, "ctrl")
  gene_dds_list[[i]] <- DESeq(gene_dds_list[[i]])
  gene_res_list[[i]] <- lfcShrink(gene_dds_list[[i]], test)
  gene_exp_list[[i]] <- getAverage(gene_dds_list[[i]])
  gene_res_list_df[[i]] <- as.data.frame(gene_res_list[[i]])
  gene_res_list_df[[i]]$gene_name <- rownames(gene_res_list_df[[i]])
  gene_vst_list[[i]] <- varianceStabilizingTransformation(gene_dds_list[[i]])
}

```

## PCA plots

All together (first 3) coloured by condition, cell type, and cell line. One per experiment, coloured by condition in each experiment.
```{r}
gene_vst <- varianceStabilizingTransformation(gene_dds)
# PCA plots 
plotPCA(gene_vst, intgroup="condition") + ggtitle("All samples: Condition") + theme_bw()
plotPCA(gene_vst, intgroup="celltype") + ggtitle("All samples: Cell type") + theme_bw()
plotPCA(gene_vst, intgroup="cellline") + ggtitle("All samples: Cell line") + theme_bw()
plotPCA(gene_vst_list$microglia.H9.IFNg, intgroup="condition") + ggtitle("Microglia H9 IFNg: Condition") + theme_bw()
plotPCA(gene_vst_list$microglia.H9.TNFa, intgroup="condition") + ggtitle("Microglia H9 TNFa: Condition") + theme_bw()
plotPCA(gene_vst_list$microglia.H9.PolyIC, intgroup="condition") + ggtitle("Microglia H9 PolyIC: Condition") + theme_bw()
plotPCA(gene_vst_list$neurons.KOLF.IFNg, intgroup="condition") + ggtitle("Neurons KOLF IFNg: Condition") + theme_bw()
plotPCA(gene_vst_list$neurons.RC17.IFNg, intgroup="condition") + ggtitle("Neurons RC17 IFNg: Condition") + theme_bw()
plotPCA(gene_vst_list$neurons.KOLF.TNFa, intgroup="condition") + ggtitle("Neurons KOLF TNFa: Condition") + theme_bw()
plotPCA(gene_vst_list$neurons.RC17.TNFa, intgroup="condition") + ggtitle("Neurons RC17 TNFa: Condition") + theme_bw()
plotPCA(gene_vst_list$neurons.KOLF.PolyIC, intgroup="condition") + ggtitle("Neurons KOLF PolyIC: Condition") + theme_bw()
plotPCA(gene_vst_list$neurons.RC17.PolyIC, intgroup="condition") + ggtitle("Neurons RC17 PolyIC: Condition") + theme_bw()
```

## Visualization

Mean plots, volcano plots and excel tables...
```{r}
# Write out the results from gene differential expression analysis
# openxlsx::write.xlsx(gene_res_list_df, file = "/Volumes/MyPassport/ASAP_invitro/results/tables/genes_DEA.xlsx")
# Create volcano and mean plots per experiment
volcano_plots_list <- list()
mean_plots_list <- list()
i <- "neurons.RC17.PolyIC"
for(i in names(gene_res_list)){
  tmp <- EnhancedVolcano(gene_res_list[[i]], 
                                             drawConnectors = T, 
                                             lab = NA, 
                                             x = 'log2FoldChange',
                                             y = 'padj',
                                             pCutoff = 0.05,
                                             FCcutoff = 1)
  tmp$data$Sig <- ifelse(tmp$data$log2FoldChange > 1 & tmp$data$Sig == "FC_P", "Upregulated", 
                                             ifelse(tmp$data$log2FoldChange < -1 & tmp$data$Sig == "FC_P", "Downregulated", "Not significant"))
  tmp$data$Sig <- paste(tmp$data$Sig, table(tmp$data$Sig)[tmp$data$Sig], sep = ": ")
  
  not_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Not significant")]
  up_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Upregulated")]
  down_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Downregulated")]
  
  point_colours <- c("grey","darkturquoise", "tomato")
  names(point_colours) <- c(not_sig_lab, down_sig_lab, up_sig_lab)
  point_sizes <- c(1,2,2)
  names(point_sizes) <- c(not_sig_lab, down_sig_lab, up_sig_lab)  
  
  title <- str_replace_all(i, pattern = "_", replacement = " ")
  
  tmp <- tmp$data
  tmp$gene_name <- rownames(tmp)
  tmp <- tmp[order(tmp$log2FoldChange, decreasing = T), ]
  tmp$gene_name <- ifelse(tmp$gene_name %in% tmp[which(startsWith(tmp$Sig, "Upregulated")),"gene_name"], tmp$gene_name, "")
  tmp$gene_name <- ifelse(tmp$gene_name %in% tmp[order(tmp$log2FoldChange, decreasing = T)[1:20],"gene_name"], tmp$gene_name, "")
  
  volcano_plots_list[[i]] <- ggplot(tmp, aes(x=log2FoldChange, y=-log10(padj), colour=Sig, size=Sig, label=gene_name)) + geom_point(alpha=0.7) + 
    geom_vline(xintercept = 1, linetype="dotted") + geom_vline(xintercept = -1, linetype="dotted") + 
    geom_hline(yintercept = -log10(0.05), linetype="dotted") + labs(colour="") + 
    geom_label_repel(max.overlaps = 100, label.size = 0.1, segment.size = 0.1, fill = alpha(c("white"),0.5)) +
    scale_color_manual(values = point_colours)  + 
    scale_size_manual(values = point_sizes, guide="none") +
    theme_bw() +
    theme(text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
    ggtitle(title)
  
  effect <- unique(samplesheet_list[[i]][which(samplesheet_list[[i]]$condition != "ctrl"),"condition"])
  mean_plots_list[[i]] <- meanPlot_cus(exp = gene_exp_list[[i]]$Mean, 
                                       test = gene_res_list[[i]], #col1 = "tomato", col2 = "grey", col3 = "tomato",
                                       c1 = effect, c2="ctrl", p = 0.05, l=1, ttl = title, repel = F, highlights = tmp$gene_name)
    
}

# Order to plot
order <- c("microglia.H9.IFNg",
           "microglia.H9.TNFa",
           "microglia.H9.PolyIC",
           "neurons.RC17.IFNg",
           "neurons.RC17.TNFa",
           "neurons.RC17.PolyIC")

volcano_plots_list[order]
mean_plots_list[order]
```

## Gene Set Enrichment Analysis

Dot plots from top terms + excel tables. Looped per experiment.

I've done this before making this R notebook, and clusterprofiler doesnt take well the seeds (although i specify it..). 

So, I leave the code here but the first time I ran things I saved them to `genes_GSEA.xlsx` which will be in the github and I'll use it here to plot the results. 
```{r}
set.seed(10) 

gse_list <- list()
for(i in names(gene_res_list_df)){
  genelist <- gene_res_list_df[[i]][,c("log2FoldChange"), drop=F]
  genelist$gene_name <- rownames(genelist)
  genelist <- genelist[order(genelist$log2FoldChange, decreasing = T),]
  genelist_FC <- genelist$log2FoldChange
  names(genelist_FC) <- genelist$gene_name
  
  gse_list[[i]] <- gseGO(geneList=sort(genelist_FC, decreasing = T), 
               ont ="BP", 
               keyType = "SYMBOL", 
               minGSSize = 3, 
               maxGSSize = 800, 
               pvalueCutoff = 0.05,
               verbose = TRUE, 
               seed = TRUE,
               OrgDb = org.Hs.eg.db, 
               pAdjustMethod = "BH")
}


# Making a list of data frames with the results
gse_list_df <- list("microglia.H9.IFNg" = gse_list$microglia.H9.IFNg@result,
                    "microglia.H9.TNFa" = gse_list$microglia.H9.TNFa@result,
                    "microglia.H9.PolyIC" = gse_list$microglia.H9.PolyIC@result,
                    "neurons.KOLF.IFNg" = gse_list$neurons.KOLF.IFNg@result,
                    "neurons.RC17.IFNg" = gse_list$neurons.RC17.IFNg@result,
                    "neurons.KOLF.TNFa" = gse_list$neurons.KOLF.TNFa@result,
                    "neurons.RC17.TNFa" = gse_list$neurons.RC17.TNFa@result,
                    "neurons.KOLF.PolyIC" = gse_list$neurons.KOLF.PolyIC@result,
                    "neurons.RC17.PolyIC" = gse_list$neurons.RC17.PolyIC@result)

# Write to excel
# openxlsx::write.xlsx(gse_list_df, "/Volumes/MyPassport/ASAP_invitro/results/tables/genes_GSEA.xlsx")
# save.image("~/Desktop/ASAP_invitro/gene_DEA.RData")

# Read excel (see description of chunk).
gse_list_df <- list()
gse_list_df[["microglia.H9.IFNg"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_invitro/results/tables/genes_GSEA.xlsx", sheet = "microglia.H9.IFNg")
gse_list_df[["microglia.H9.TNFa"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_invitro/results/tables/genes_GSEA.xlsx", sheet = "microglia.H9.TNFa")
gse_list_df[["microglia.H9.PolyIC"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_invitro/results/tables/genes_GSEA.xlsx", sheet = "microglia.H9.PolyIC")
gse_list_df[["neurons.RC17.IFNg"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_invitro/results/tables/genes_GSEA.xlsx", sheet = "neurons.RC17.IFNg")
gse_list_df[["neurons.RC17.TNFa"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_invitro/results/tables/genes_GSEA.xlsx", sheet = "neurons.RC17.TNFa")
gse_list_df[["neurons.RC17.PolyIC"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_invitro/results/tables/genes_GSEA.xlsx", sheet ="neurons.RC17.PolyIC")

# Handy function to plot only top 20 most enriched terms
top20_gsea_dotplot <- function(gse, name){
  gse@result$sign <- ifelse(gse@result$enrichmentScore > 0, "Activated", "Supressed")
  gse@result$num_genes <- sapply(str_split(gse@result$core_enrichment, "/"), length)
  gse@result$gene_ratio <- gse@result$num_genes / gse@result$setSize
  
  gse@result <- top_n(gse@result, -20, gse@result$p.adjust) %>% arrange(gene_ratio)
  if(nrow(gse@result) > 20){
    gse@result <- top_n(gse@result, 20, abs(gse@result$enrichmentScore)) %>% arrange(gene_ratio)
  }
  gse@result$Description <- factor(gse@result$Description, levels = unique(gse@result$Description))
  
  ggplot(gse@result, aes(x=Description, y=gene_ratio, size = num_genes, fill = p.adjust)) + geom_point(shape=21, colour="lightgrey") + facet_wrap(.~sign) + coord_flip() + theme_pubr(legend = "right", border = T) + labs(x="", fill="Padj", y = "Gene ratio", size = "Num genes") + scale_fill_gradientn(colors = c("firebrick1", "gray94")) + scale_size_continuous(range = c(2,8)) + 
    theme(axis.text.y = element_text(size=10), strip.text.x = element_text(size = 10),axis.text.x = element_text(size=10),legend.text = element_text(size=10),legend.title = element_text(size=10)) + 
    ggtitle(paste(name, ":\nTop 20 most significant"))
}

# Loop again one plot per experiment
top20_gsea <- list()
for( g in names(gse_list)){
  print(g)
  print(nrow(gse_list[[g]]@result))
  if(nrow(gse_list[[g]]@result) > 0){
    top20_gsea[[g]] <- top20_gsea_dotplot(gse = gse_list[[g]], name=g)  
  }
}

# Aaaand plot!
top20_gsea$microglia.H9.IFNg
top20_gsea$neurons.KOLF.IFNg
top20_gsea$neurons.RC17.IFNg
top20_gsea$microglia.H9.PolyIC
top20_gsea$neurons.KOLF.PolyIC
top20_gsea$neurons.RC17.PolyIC
top20_gsea$microglia.H9.TNFa
top20_gsea$neurons.KOLF.TNFa
top20_gsea$neurons.RC17.TNFa

```

## Another way of visualizing the results (barplots, top 10)
```{r}
direction_cols <- c("Activated" = "tomato", "Supressed" = "darkturquoise")

gse_list_df$microglia.H9.IFNg %>% 
  arrange(p.adjust, -enrichmentScore) %>%
  slice(1:10) %>% 
  mutate(direction = ifelse(enrichmentScore > 0, "Activated", "Supressed")) %>% 
  ggplot(aes(x=enrichmentScore, y=Description, fill=direction)) + geom_bar(stat="identity") + 
  theme_bw() + labs(x="Enrichment Score", y="")+ 
  ggtitle("microglia.H9.IFNg") + scale_fill_manual(values = direction_cols)

gse_list_df$microglia.H9.TNFa %>% 
  arrange(p.adjust, -enrichmentScore) %>%
  slice(1:10) %>% 
  mutate(direction = ifelse(enrichmentScore > 0, "Activated", "Supressed")) %>% 
  ggplot(aes(x=enrichmentScore, y=Description, fill=direction)) + geom_bar(stat="identity") + 
  theme_bw() + labs(x="Enrichment Score", y="") + 
  ggtitle("microglia.H9.TNFa") + scale_fill_manual(values = direction_cols)

gse_list_df$microglia.H9.PolyIC %>% 
  arrange(p.adjust, -enrichmentScore) %>%
  slice(1:10) %>% 
  mutate(direction = ifelse(enrichmentScore > 0, "Activated", "Supressed")) %>% 
  ggplot(aes(x=enrichmentScore, y=Description, fill=direction)) + geom_bar(stat="identity") + 
  theme_bw() + labs(x="Enrichment Score", y="")+ 
  ggtitle("microglia.H9.PolyIC") + scale_fill_manual(values = direction_cols)

gse_list_df$neurons.RC17.IFNg %>% 
  arrange(p.adjust, -enrichmentScore) %>%
  slice(1:10) %>% 
  mutate(direction = ifelse(enrichmentScore > 0, "Activated", "Supressed")) %>% 
  ggplot(aes(x=enrichmentScore, y=Description, fill=direction)) + geom_bar(stat="identity") + 
  theme_bw() + labs(x="Enrichment Score", y="")+ 
  ggtitle("neurons.RC17.IFNg") + scale_fill_manual(values = direction_cols)

gse_list_df$neurons.RC17.TNFa %>% 
  arrange(p.adjust, -enrichmentScore) %>%
  slice(1:10) %>% 
  mutate(direction = ifelse(enrichmentScore > 0, "Activated", "Supressed")) %>% 
  ggplot(aes(x=enrichmentScore, y=Description, fill=direction)) + geom_bar(stat="identity") + 
  theme_bw() + labs(x="Enrichment Score", y="")+ 
  ggtitle("neurons.RC17.TNFa") + scale_fill_manual(values = direction_cols)

gse_list_df$neurons.RC17.PolyIC %>% 
  arrange(p.adjust, -enrichmentScore) %>%
  slice(1:10) %>% 
  mutate(direction = ifelse(enrichmentScore > 0, "Activated", "Supressed")) %>% 
  ggplot(aes(x=enrichmentScore, y=Description, fill=direction)) + geom_bar(stat="identity") + 
  theme_bw() + labs(x="Enrichment Score", y="") + 
  ggtitle("neurons.RC17.PolyIC") + scale_fill_manual(values = direction_cols)

```

## How well does the two cell lines correlate? (Neurons)
```{r}
# Taking the effect (log2FC) of all genes in the result objects (one per experiment). If the cell lines replicate each other, you should see similar effects 
gene_res_neurons <- data.frame("gene" = rownames(gene_res_list_df$microglia.H9.IFNg),
                               "log2FC_neurons.RC17.IFNg" = gene_res_list_df$neurons.RC17.IFNg[rownames(gene_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.KOLF.IFNg" = gene_res_list_df$neurons.KOLF.IFNg[rownames(gene_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.RC17.PolyIC" = gene_res_list_df$neurons.RC17.PolyIC[rownames(gene_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.KOLF.PolyIC" = gene_res_list_df$neurons.KOLF.PolyIC[rownames(gene_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.RC17.TNFa" = gene_res_list_df$neurons.RC17.TNFa[rownames(gene_res_list_df$microglia.H9.IFNg), "log2FoldChange"],
                               "log2FC_neurons.KOLF.TNFa" = gene_res_list_df$neurons.KOLF.TNFa[rownames(gene_res_list_df$microglia.H9.IFNg), "log2FoldChange"])


# Plot out!
ggarrange(
  gene_res_neurons %>% 
  drop_na(log2FC_neurons.RC17.IFNg, log2FC_neurons.KOLF.IFNg) %>% 
ggplot(aes(x=log2FC_neurons.RC17.IFNg, y=log2FC_neurons.KOLF.IFNg)) + 
  geom_point(alpha=0.5) + theme_bw() + 
  geom_hline(yintercept = 0, linetype="dashed", color="lightgrey") + 
  geom_vline(xintercept = 0, linetype="dashed", color="lightgrey") + 
  labs(y = "log2FC KOLF IFNg", x = "log2FC RC17 IFNg") + geom_smooth(method = "lm", color="darkolivegreen2") + # Linear model to correlate 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) + ggtitle("Neurons + IFNg"), # R2 and p value

gene_res_neurons %>% 
  drop_na(log2FC_neurons.RC17.TNFa, log2FC_neurons.KOLF.TNFa) %>% 
ggplot(aes(x=log2FC_neurons.RC17.TNFa, y=log2FC_neurons.KOLF.TNFa)) + 
  geom_point(alpha=0.5) + theme_bw() + 
  geom_hline(yintercept = 0, linetype="dashed", color="lightgrey") + 
  geom_vline(xintercept = 0, linetype="dashed", color="lightgrey") + 
  labs(y = "log2FC KOLF TNFa", x = "log2FC RC17 TNFa") + geom_smooth(method = "lm", color="darkolivegreen2") + 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) + ggtitle("Neurons + TNFa"),

gene_res_neurons %>% 
  drop_na(log2FC_neurons.RC17.PolyIC, log2FC_neurons.KOLF.PolyIC) %>% 
ggplot(aes(x=log2FC_neurons.RC17.PolyIC, y=log2FC_neurons.KOLF.PolyIC)) + 
  geom_point(alpha=0.5) + theme_bw() + 
  geom_hline(yintercept = 0, linetype="dashed", color="lightgrey") + 
  geom_vline(xintercept = 0, linetype="dashed", color="lightgrey") + 
  labs(y = "log2FC KOLF PolyIC", x = "log2FC RC17 PolyIC") + geom_smooth(method = "lm", color="darkolivegreen2") + 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) + ggtitle("Neurons + PolyIC"))

```

## What is the overlap of gene effects between the three treatments?
```{r}
upset(fromList(list("microglia.H9.IFNg" = rownames(gene_res_list$microglia.H9.IFNg[which(gene_res_list$microglia.H9.IFNg$log2FoldChange > 1 & gene_res_list$microglia.H9.IFNg$padj < 0.05),]),
"microglia.H9.PolyIC" = rownames(gene_res_list$microglia.H9.PolyIC[which(gene_res_list$microglia.H9.PolyIC$log2FoldChange > 1 & gene_res_list$microglia.H9.PolyIC$padj < 0.05),]),
"microglia.H9.TNFa" = rownames(gene_res_list$microglia.H9.TNFa[which(gene_res_list$microglia.H9.TNFa$log2FoldChange > 1 & gene_res_list$microglia.H9.TNFa$padj < 0.05),]))))

upset(fromList(list("microglia.H9.IFNg" = rownames(gene_res_list$microglia.H9.IFNg[which(gene_res_list$microglia.H9.IFNg$log2FoldChange < -1 & gene_res_list$microglia.H9.IFNg$padj < 0.05),]),
"microglia.H9.PolyIC" = rownames(gene_res_list$microglia.H9.PolyIC[which(gene_res_list$microglia.H9.PolyIC$log2FoldChange < -1 & gene_res_list$microglia.H9.PolyIC$padj < 0.05),]),
"microglia.H9.TNFa" = rownames(gene_res_list$microglia.H9.TNFa[which(gene_res_list$microglia.H9.TNFa$log2FoldChange < -1 & gene_res_list$microglia.H9.TNFa$padj < 0.05),]))))
```

## As an example, JAK, STAT, and CIITA
```{r}
jakstat <- as.data.frame(gene_count_norm[c("JAK2", "STAT1", "CIITA"),]) 
jakstat$gene_name <- rownames(jakstat)
samplesheet$name <- samplesheet$sample

jakstat %>% 
  pivot_longer(!gene_name) %>% 
  left_join(samplesheet) %>% 
  filter(cellline == "H9") %>% 
  filter(celltype == "microglia") %>% 
  filter(condition %in% c("IFNg", "PolyIC", "TNFa", "ctrl")) %>% 
  ggplot(aes(x=condition, y=log2(value), fill = condition)) + 
  geom_jitter(width = 0.2, aes(color=condition)) + geom_boxplot(alpha=0.5) + 
  facet_wrap(.~gene_name, scales = "free") + theme_bw() + labs(y="log2(Normalized Expression)", x="") 
```

