---
title: "Microglia Gene Expression Analysis: batch and sex covariates"
author: "Raquel Garza"
output:
  html_document:
    df_print: paged
---
## Overview

This notebook is a repetition of gene_DEA_GSEA.Rmd to verify that the results hold when adjusting for batch and sex effects in the design formula of the differential expression analyses.

### Inputs
- trusTEr output files with gene counts per cluster/region
- Sample metadata: ASAP_samplesheet.xlsx

### Outputs
- Differential expression results in microglia (excel files)
- Gene set enrichment analyses results (excel files)
- Figures showing:
  - PD vs Control effects
  - IFN enrichment across clusters

### Notes
- Only expressed genes (row sums > 0) are tested.

## Setup: Load libraries and helper functions
Load all required libraries and custom DEA functions
```{r class.source = 'fold-hide'}
library(DESeq2)
library(data.table)
library(ggpubr)
library(ggplot2)
library(tidyverse)
library(openxlsx)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(EnhancedVolcano)


source("TE_DEA_functions.R")
```

## Characterize PD effect per cluster
Differential expression analyses for PD vs Control pseudobulks (adjusting for batch and sex) for each region in microglia (cluster 4). 
Save results (one excel with region per sheet).
```{r}
samplesheet <- read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/ASAP_samplesheet.xlsx")
gene_dds_list <- list()
gene_exp_list <- list()
gene_res_list <- list()
gene_res_list_df <- list()
regions <- c("SN", "PUT", "AMY", "PFC")
cluster <- "4" # Microglia cluster 
gene_dds_list[[cluster]] <- list()
gene_exp_list[[cluster]] <- list()
gene_res_list[[cluster]] <- list()
gene_res_list_df[[cluster]] <- list()
for(region in regions){
  print(cluster)
  print(region)
  gene_counts <- list()
  gene_counts[[region]] <- fread(paste("~/inbox/ASAP/trusTEr_output_vs2/", region,"/gene_counts_", region,"_",cluster,".csv", sep=""), sep = ",", data.table = F)
  rownames(gene_counts[[region]]) <- gene_counts[[region]]$Geneid

  samplesheet_list <- split(samplesheet, f = samplesheet$Region)

  rownames(samplesheet_list[[region]]) <- samplesheet_list[[region]]$Sample
  samples_cluster <- paste(rownames(samplesheet_list[[region]]), cluster, sep="_")
  samples_cluster <- samples_cluster[which(samples_cluster %in% colnames(gene_counts[[region]]))]
  rownames(samplesheet_list[[region]]) <- paste(rownames(samplesheet_list[[region]]), cluster, sep="_")
  rownames(gene_counts[[region]]) <- gene_counts[[region]]$Geneid
  
  expressed_genes <- rownames(gene_counts[[region]])[which(rowSums(gene_counts[[region]][,samples_cluster]) > 0)]
  gene_dds_list[[cluster]][[region]] <- DESeqDataSetFromMatrix((gene_counts[[region]][expressed_genes,samples_cluster]+1), samplesheet_list[[region]][samples_cluster,], design =  ~ Seqnum + Sex + Dx) # Covariates with batch and sex
  gene_dds_list[[cluster]][[region]]$Dx <- relevel(gene_dds_list[[cluster]][[region]]$Dx, "Ctl")
  gene_dds_list[[cluster]][[region]] <- DESeq(gene_dds_list[[cluster]][[region]])
  gene_res_list[[cluster]][[region]] <- results(gene_dds_list[[cluster]][[region]], )
  gene_res_list_df[[cluster]][[region]] <- as.data.frame(gene_res_list[[cluster]][[region]])
  gene_res_list_df[[cluster]][[region]]$gene_id <- rownames(gene_res_list_df[[cluster]][[region]])
  print(names(gene_res_list_df[[cluster]]))
}
openxlsx::write.xlsx(gene_res_list_df[[cluster]], paste("/Volumes/MyPassport/ASAP/data/ASAP_PMDBS_snRNAseq/results/tables/gene_cluster_", cluster, "_PD_DEA_batch_sex_snRNAseq_res.xlsx", sep=""))


```

## Some helper functions for Gene Set Enrichment Analyses
```{r}
set.seed(10)
gse_dotplot <- function(df){
  genelist <- df[which(!is.na(df$log2FoldChange)),c("log2FoldChange", "gene_id"), drop=F]
  genelist <- genelist[order(genelist$log2FoldChange, decreasing = T),]

  genelist_FC <- genelist$log2FoldChange
  names(genelist_FC) <- genelist$gene_id
  gse <- gseGO(geneList=genelist_FC,
               ont ="ALL",
               keyType = "SYMBOL",
               minGSSize = 3,
               maxGSSize = 800,
               seed = T,
               pvalueCutoff = 0.05,
               verbose = TRUE,
               OrgDb = org.Hs.eg.db,
               pAdjustMethod = "BH")
  return(gse)
}

gse_pretty_dotplot <- function(gse, topn = 10, terms = NA){
  gse <- gse@result
  # gse <- gse[which(gse$ONTOLOGY == "BP"),]
  gse$sign <- ifelse(gse$enrichmentScore > 0, "Activated", "Supressed")
  gse$num_genes <- sapply(str_split(gse$core_enrichment, "/"), length)
  gse$gene_ratio <- gse$num_genes / gse$setSize
  gse$Description <- factor(gse$Description, levels = gse[order(gse$gene_ratio), "Description"])
  
  if(!all(is.na(terms))){
    gse <- gse %>% 
      drop_na() %>% 
      filter(ID %in% terms)
  }else{
    gse <- gse[order(gse$NES, -log10(gse$p.adjust), decreasing = T),]
  }
  gse %>% 
      drop_na() %>% 
      ggplot(aes(x=Description, y=gene_ratio, size = num_genes, fill = p.adjust)) + geom_point(shape=21, colour="lightgrey") + facet_wrap(.~sign) + coord_flip() + theme_pubr(legend = "right", border = T) + labs(x="", fill="Padj", y = "Gene ratio", size = "Num genes") + scale_fill_gradientn(colors = c("firebrick1", "gray94")) + scale_size_continuous(range = c(2,8)) + theme(axis.text.y = element_text(size=10), strip.text.x = element_text(size = 10),axis.text.x = element_text(size=10),legend.text = element_text(size=10),legend.title = element_text(size=10)) 
}
```

## Gene Set Enrichment Analyses (GSEA)

For some reason the seed is not taken in by clusterProfiler, so the results from here onwards will be using the original run I did on this code (same but ever so slightly different numbers, loaded here at the end of the chunk). The clean data provided will be the original run I did. 

* Test whether there is enrichment of GO-terms within the DEA results (log2FC based).
* Save results to excel 
```{r}
gse_list <- list()
gse_list_df <- list()
gse_plot_list <- list()
gse_list[[cluster]] <- list()
gse_list_df[[cluster]] <- list()
gse_plot_list[[cluster]] <- list()
for(region in regions){ 
  print(cluster)
  print(region)
  gse_list[[cluster]][[region]] <- gse_dotplot(df = gene_res_list_df[[cluster]][[region]])
  gse_list_df[[cluster]][[region]] <- gse_list[[cluster]][[region]]@result
  if(!is.null(gse_plot_list[[cluster]][[region]])){
  # gse_plot_list[[cluster]][[region]] <- gse_pretty_dotplot(gse = gse_list[[cluster]][[region]], topn = 30) + ggtitle(paste(region, cluster, sep=" cluster: "))
  }
}
if(length(gse_list_df[[cluster]]) > 0){
openxlsx::write.xlsx(gse_list_df[[cluster]], paste("/Volumes/MyPassport/ASAP/data/ASAP_PMDBS_snRNAseq/results/tables/gse_cluster", cluster, "PD_GSEA_batch_sex_snRNAseq_res.xlsx", sep="_"))
}

```

## IFN and viral related terms in microglia 
```{r}
cluster = "4"
region = "SN"
ifn_viral_related_go <- gse_list_df[[cluster]][[region]][which(grepl("interferon|vir", gse_list_df[[cluster]][[region]]$Description)), "ID"]
gse_pretty_dotplot(gse = gse_list[[cluster]][[region]], topn = 10, terms = c(ifn_viral_related_go)) + ggtitle(paste(region, cluster, sep=" cluster: "))

region = "PUT"
ifn_viral_related_go <- gse_list_df[[cluster]][[region]][which(grepl("interferon|vir", gse_list_df[[cluster]][[region]]$Description)), "ID"]
gse_pretty_dotplot(gse = gse_list[[cluster]][[region]], topn = 10, terms = c(ifn_viral_related_go)) + ggtitle(paste(region, cluster, sep=" cluster: "))

region = "AMY"
ifn_viral_related_go <- gse_list_df[[cluster]][[region]][which(grepl("interferon|vir", gse_list_df[[cluster]][[region]]$Description)), "ID"]
gse_pretty_dotplot(gse = gse_list[[cluster]][[region]], topn = 10, terms = c(ifn_viral_related_go)) + ggtitle(paste(region, cluster, sep=" cluster: "))

region = "PFC"
ifn_viral_related_go <- gse_list_df[[cluster]][[region]][which(grepl("interferon|vir", gse_list_df[[cluster]][[region]]$Description)), "ID"]
ifn_viral_related_go
```

## IFN related genes avg log2FC heatmaps

IFN related genes that were up in PD SN microglia
```{r}
gene_res_list_df$`4`$SN$region <- "SN"
gene_res_list_df$`4`$PUT$region <- "PUT"
gene_res_list_df$`4`$PFC$region <- "PFC"
gene_res_list_df$`4`$AMY$region <- "AMY"

ifn_related_go <- gse_list_df[["4"]][["SN"]][which(grepl("interferon", gse_list_df[["4"]][["SN"]]$Description)), "Description"]
tmp <- sapply(sapply(gse_list_df[["4"]][["SN"]][which(gse_list_df[["4"]][["SN"]]$Description %in% ifn_related_go), "core_enrichment"], str_split, "/"), unlist)
names(tmp) <- NULL
tmp <- unique(unlist(tmp))

tmp_res <- rbind(gene_res_list_df$`4`$SN[tmp,],
gene_res_list_df$`4`$PUT[tmp,],
gene_res_list_df$`4`$PFC[tmp,],
gene_res_list_df$`4`$AMY[tmp,])

tmp_res <- unique(tmp_res)

tmp_heatmap_log2FC <- reshape2::dcast(tmp_res, gene_id~region, value.var = "log2FoldChange")
rownames(tmp_heatmap_log2FC) <- tmp_heatmap_log2FC$gene_id
tmp_heatmap_pvalue <- reshape2::dcast(tmp_res, gene_id~region, value.var = "pvalue")
tmp_heatmap_pvalue[is.na(tmp_heatmap_pvalue)] <- 1
rownames(tmp_heatmap_pvalue) <- tmp_heatmap_pvalue$gene_id
tmp_heatmap_pvalue <- tmp_heatmap_pvalue[,-1]
tmp_heatmap_pvalue <- ifelse(tmp_heatmap_pvalue > 0.05, "", ifelse(tmp_heatmap_pvalue < 0.001, "***", ifelse(tmp_heatmap_pvalue < 0.01, "**", "*")))

bk_neg <- seq(-3,0, length.out=50)
bk_pos <- seq(0,3, length.out=50)
bk <- c(bk_neg, bk_pos)
bk <- unique(bk)
color_pos <- colorRampPalette(c("white", "red"))(50)
color_neg <- colorRampPalette(c("blue", "white"))(50)
tmp_heatmap_log2FC[is.na(tmp_heatmap_log2FC)] <- 0

pheatmap::pheatmap(tmp_heatmap_log2FC[,-1], breaks = bk, color = c(color_neg, color_pos), cluster_cols = F, main = "Core enrichment of enriched IFN-related\nGO terms in SN microglia", show_rownames = T, border_color = F, display_numbers = tmp_heatmap_pvalue)

pheatmap::pheatmap(tmp_heatmap_log2FC[,-1], breaks = bk, color = c(color_neg, color_pos), cluster_cols = F, main = "Core enrichment of enriched IFN-related\nGO terms in SN microglia", show_rownames = F, border_color = F, display_numbers = tmp_heatmap_pvalue)

```
