---
title: "Microglia ERV Differential Expression Analysis: batch and sex covariates"
author: "Raquel Garza"
output:
  html_document:
    df_print: paged
---

## Overview 

This notebook is a repetition of HERV_expression_PDvsCtl.Rmd to verify that the results hold when including genes and TEs into the same data frame and adjusting for batch and sex effects in the design formula of the differential expression analyses.

## Set up
Loading data and needed libraries
```{r}
library(DESeq2)
library(data.table)
library(ggpubr)
library(EnhancedVolcano)
library(limma)
library(variancePartition)
library(ggplot2)
library(tidyverse)
library(ggpmisc)
library(openxlsx)
source("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/processing/TE_DEA_functions.R")
source("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/processing/TE_heatmap_functions.R")
# Load samplesheet
samplesheet <- read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/ASAP_samplesheet.xlsx")
regions <- c("SN", "PUT", "PFC", "AMY")
# Load ids of expressed elements in bulk data
expressed_hervs_bulkrna <- fread("/Volumes/MyPassport/ASAP/data/ASAP_PMDBS_bulkRNAseq/results/tables/expressed_ervs_expressed.tab", data.table = F, header = F)$V1
expressed_l1s_bulkrna <- fread("/Volumes/MyPassport/ASAP/data/ASAP_PMDBS_bulkRNAseq/results/tables/expressed_FL_L1HS_L1PA3_expressed.tab", data.table = F, header = F)$V1
```

## 1. Differential Expression Analysis (DEA) with DESeq2

* Build DESeq2 datasets for each region-cluster combination
* Adjust for batch and sex
* Calculate variance per covariate using limma voom
* Compare PD vs. Control, apply shrinkage (lfcShrink)
* Save results as excel tables (clean data)

```{r}
te_dds_list <- list()
te_alt_dds_list <- list()
te_exp_list <- list()
te_alt_exp_list <- list()
te_res_list <- list()
te_alt_res_list <- list()
te_res_list_df <- list()
te_alt_res_list_df <- list()
te_varPart_list <- list()
te_var_plot_list <- list()
regions <- c("SN", "PUT", "PFC", "AMY")
cluster = "4"
te_dds_list[[cluster]] <- list()
te_alt_dds_list[[cluster]] <- list()
te_exp_list[[cluster]] <- list()
te_alt_exp_list[[cluster]] <- list()
te_res_list[[cluster]] <- list()
te_alt_res_list[[cluster]] <- list()
te_res_list_df[[cluster]] <- list()
te_alt_res_list_df[[cluster]] <- list()
for(region in regions){
  print(cluster)
  print(region)
  te_counts <- list()
  gene_te_counts <- fread(paste("~/inbox/ASAP/trusTEr_output_vs2/", region,"/gene_counts_", region,"_",cluster,".csv", sep=""), sep = ",", data.table = F)
  hervs_te_counts <- fread(paste("~/inbox/ASAP/trusTEr_output_vs2/", region,"/HERV_counts_", region,"_",cluster,".csv", sep=""), sep = ",", data.table = F)
  rownames(hervs_te_counts) <- hervs_te_counts$Geneid
  hervs_te_counts <- hervs_te_counts[expressed_hervs_bulkrna,]
  l1s_te_counts <- fread(paste("~/inbox/ASAP/trusTEr_output_vs2/", region,"/L1_expressed_counts_", region,"_",cluster,".csv", sep=""), data.table = F)
  te_counts[[region]] <- rbind(gene_te_counts, l1s_te_counts, hervs_te_counts)
  rownames(te_counts[[region]]) <- te_counts[[region]]$Geneid
  
  samplesheet_list <- split(samplesheet, f = samplesheet$Region)
  num_nuclei <- read.table(paste("/Volumes/MyPassport/ASAP/data/ASAP_PMDBS_snRNAseq/results/tables/number_nuclei/",region, "_clusters.txt", sep=""), header = F, skip = 1)
  colnames(num_nuclei) <- c("cluster", "num_nuclei")
  # There is a "total" row. I'll remove that for convenience
  num_nuclei <- num_nuclei[which(num_nuclei$cluster != "total"),]
  rownames(num_nuclei) <- num_nuclei$cluster
  # Just keep pseudobulks with at least 100 nuclei (arbitrary)
  num_nuclei <- num_nuclei[which(num_nuclei$num_nuclei > 100),] # Remove tiny pseudobulks
  
  rownames(samplesheet_list[[region]]) <- samplesheet_list[[region]]$Sample
  samples_cluster <- paste(rownames(samplesheet_list[[region]]), cluster, sep="_")
  samples_cluster <- samples_cluster[which(samples_cluster %in% colnames(te_counts[[region]])  & samples_cluster %in% num_nuclei$cluster)]
  rownames(samplesheet_list[[region]]) <- paste(rownames(samplesheet_list[[region]]), cluster, sep="_")
  rownames(te_counts[[region]]) <- te_counts[[region]]$Geneid
  # Expressed in bulk RNA
  
  if(length(samples_cluster) > 0){
    if(all(table(samplesheet_list[[region]][samples_cluster, "Dx"]) > 3)){
      # Expressed in snRNA
      te_counts[[region]] <- te_counts[[region]][which(rowSums(te_counts[[region]][,samples_cluster]) > 0),]
      
      # Assess covariate variance
      design <- model.matrix(~ Seqnum + Sex + Dx, data = samplesheet_list[[region]][samples_cluster,])
      form <- ~ Seqnum + Sex + Dx
      vobj <- voom(te_counts[[region]][,samples_cluster]+1, design)
      region_cluster <- paste(region, cluster, sep="_")
      tryCatch({
        te_varPart_list[[region_cluster]] <- fitExtractVarPartModel(vobj, form, samplesheet_list[[region]][samples_cluster,])
        te_var_plot_list[[region_cluster]] <- plotVarPart(te_varPart_list[[region_cluster]]) + geom_hline(yintercept = 10)    
        te_varPart_list[[region_cluster]]$group <- region_cluster
      }, error = function(e2) {
        warning(sprintf("Failed for Cluster %s, Region %s: %s", cluster, region, e2$message))
        return(NULL)
      })
      
      if(length(samples_cluster) > 0){
        if(all(table(samplesheet_list[[region]][samples_cluster, "Dx"]) > 3)){
          te_dds_list[[cluster]][[region]] <- DESeqDataSetFromMatrix((te_counts[[region]][,samples_cluster]+1), samplesheet_list[[region]][samples_cluster,], design =  ~ Seqnum + Sex + Dx)
          te_dds_list[[cluster]][[region]]$Dx <- relevel(te_dds_list[[cluster]][[region]]$Dx, "Ctl")
          te_dds_list[[cluster]][[region]] <- DESeq(te_dds_list[[cluster]][[region]])
          te_exp_list[[cluster]][[region]] <- getAverage(te_dds_list[[cluster]][[region]])
          te_res_list[[cluster]][[region]] <- results(te_dds_list[[cluster]][[region]], )
          te_res_list_df[[cluster]][[region]] <- as.data.frame(te_res_list[[cluster]][[region]])
          te_res_list_df[[cluster]][[region]]$Geneid <- rownames(te_res_list_df[[cluster]][[region]])
          
          alt_formula <- "~"
          if(mean(te_varPart_list[[region_cluster]]$Seqnum) > 0.15){
            alt_formula <- ifelse(alt_formula == "~", paste(alt_formula, "Seqnum"), paste(alt_formula, "Seqnum",sep=" + "))
          }
          if(mean(te_varPart_list[[region_cluster]]$Sex) > 0.15){
            alt_formula <- ifelse(alt_formula == "~", paste(alt_formula, "Sex"), paste(alt_formula, "Sex",sep=" + "))
          }
          alt_formula <- paste(alt_formula, "Dx", sep=" + ")
          if(alt_formula != "~ Seqnum + Sex + Dx"){
            alt_formula <- as.formula(alt_formula)
            te_alt_dds_list[[cluster]][[region]] <- DESeqDataSetFromMatrix((te_counts[[region]][,samples_cluster]+1), samplesheet_list[[region]][samples_cluster,], design =  alt_formula)
            te_alt_dds_list[[cluster]][[region]]$Dx <- relevel(te_alt_dds_list[[cluster]][[region]]$Dx, "Ctl")
            te_alt_dds_list[[cluster]][[region]] <- DESeq(te_alt_dds_list[[cluster]][[region]])
            te_alt_exp_list[[cluster]][[region]] <- getAverage(te_alt_dds_list[[cluster]][[region]])
            te_alt_res_list[[cluster]][[region]] <- results(te_alt_dds_list[[cluster]][[region]], )
            te_alt_res_list_df[[cluster]][[region]] <- as.data.frame(te_alt_res_list[[cluster]][[region]])
            te_alt_res_list_df[[cluster]][[region]]$Geneid <- rownames(te_alt_res_list_df[[cluster]][[region]])
          }
        }
      }
    }
  }
}

do.call(rbind, te_varPart_list) %>% 
  group_by(group) %>% 
  summarize(mean_batch_var = mean(Seqnum), mean_sex_var = mean(Sex)) %>% 
  openxlsx::write.xlsx("/Volumes/MyPassport/ASAP/data/ASAP_PMDBS_snRNAseq/results/tables/mean_covariate_variance_per_pseudobulk.xlsx")
```

```{r}
get_subset_plots <- function(coldata, exp, res, title, select){
   select <- select[which(select %in% rownames(res))]
    tmp <- EnhancedVolcano(res[select,], drawConnectors = T, lab = NA, x = 'log2FoldChange',y = 'padj',pCutoff = 0.05,FCcutoff = 1)
      tmp$data$Sig <- ifelse(tmp$data$log2FoldChange > 1 & tmp$data$Sig == "FC_P", "Upregulated", 
                             ifelse(tmp$data$log2FoldChange < -1 & tmp$data$Sig == "FC_P", "Downregulated", "Not significant"))
      tmp$data$Sig <- paste(tmp$data$Sig, table(tmp$data$Sig)[tmp$data$Sig], sep = ": ")
      
      not_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Not significant")]
      up_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Upregulated")]
      down_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Downregulated")]
      
      if(length(unique(tmp$data$Sig)) == 1){
        point_colours <- c("grey")
      }
      if(length(unique(tmp$data$Sig)) == 2){
        if(any(str_detect(unique(tmp$data$Sig), "Upregulated"))){
          point_colours <- c("grey","tomato")
        }
      }
      if(length(unique(tmp$data$Sig)) == 2){
        if(any(str_detect(unique(tmp$data$Sig), "Downregulated"))){
          point_colours <- c("grey","darkturquoise")
        }
      }
      if(length(unique(tmp$data$Sig)) == 3){
        point_colours <- c("grey","darkturquoise", "tomato")
        names(point_colours) <- c(not_sig_lab, down_sig_lab, up_sig_lab)
      }
      point_sizes <- c(1,2,2)
      names(point_sizes) <- c(not_sig_lab, down_sig_lab, up_sig_lab)  
      
      tmp <- tmp$data
      volcano_plot <- ggplot(tmp, aes(x=log2FoldChange, y=-log10(padj), colour=Sig, size=Sig)) + geom_point(alpha=0.7) + 
        geom_vline(xintercept = 1, linetype="dotted") + geom_vline(xintercept = -1, linetype="dotted") + 
        geom_hline(yintercept = -log10(0.1), linetype="dotted") + labs(colour="") + 
        scale_color_manual(values = point_colours)  + 
        scale_size_manual(values = point_sizes, guide="none") +
        theme_bw() +
        theme(text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
        ggtitle(title)
      
      effect <- unique(coldata[which(coldata$Dx != "Ctl"),"Dx"])
      mean_plot <- meanPlot_cus(exp = exp$Mean[select,],
                                                           test = res, 
                                                           col1 = ifelse(length(unique(tmp$Sig)) == 3, "tomato", 
                                                                         ifelse(any(str_detect(unique(tmp$Sig), "Downregulated")), "grey", "darkturquoise")), 
                                                           col2 = ifelse(length(unique(tmp$Sig)) == 3, "darkturquoise", 
                                                                         ifelse(any(str_detect(unique(tmp$Sig), "Downregulated")), "darkturquoise", "grey")), 
                                                           col3 = ifelse(length(unique(tmp$Sig)) == 3, "grey", 
                                                                         ifelse(any(str_detect(unique(tmp$Sig), "Upregulated")), "tomato", "grey")),
                                                           c1 = effect, c2="Ctl", p = 0.05, l=1, ttl = title, repel = F)
      return(list("volcano" = volcano_plot, "mean" = mean_plot))
}
```


```{r}
herv_plots_list <- list()
l1_plots_list <- list()
genes_plots_list <- list()
all_plots_list <- list()
for(cluster in as.character(0:7)){
  for(region in regions){ 
    if(!all(is.na(te_res_list[[cluster]][[region]]))){
        coldata = samplesheet_list[[region]]
        exp = te_exp_list[[cluster]][[region]]
        res = te_res_list[[cluster]][[region]]
        title = paste(region, cluster, sep=": cluster ")
        herv_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, expressed_hervs_bulkrna)
        l1_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, expressed_l1s_bulkrna)
        genes <- rownames(res)[which(!rownames(res) %in% c(expressed_hervs_bulkrna, expressed_l1s_bulkrna))]
        genes_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, genes)
        all_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, rownames(res))
    }
  }
}

herv_alt_plots_list <- list()
l1_alt_plots_list <- list()
genes_alt_plots_list <- list()
all_alt_plots_list <- list()
for(cluster in as.character(0:7)){
  for(region in regions){ 
    if(!all(is.na(te_alt_res_list[[cluster]][[region]]))){
        coldata = samplesheet_list[[region]]
        exp = te_alt_exp_list[[cluster]][[region]]
        res = te_alt_res_list[[cluster]][[region]]
        title = paste(region, cluster, sep=": cluster ")
        herv_alt_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, expressed_hervs_bulkrna)
        l1_alt_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, expressed_l1s_bulkrna)
        genes <- rownames(res)[which(!rownames(res) %in% c(expressed_hervs_bulkrna, expressed_l1s_bulkrna))]
        genes_alt_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, genes)
        all_alt_plots_list[[title]] <- get_subset_plots(coldata, exp, res, title, rownames(res))
    }
  }
}
herv_plots_list$`SN: cluster 4`
l1_plots_list$`SN: cluster 4`

```

## How well does this correlate without batch and sex in the model? 

### ERVs
```{r}
te_res_nocov_list_df <- list()
te_res_nocov_list_df[["4"]] <- list()
te_res_nocov_list_df[["4"]][["SN"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/HERVs_cluster_4_DEA_snRNAseq.xlsx", sheet = "SN")
te_res_nocov_list_df[["4"]][["PUT"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/HERVs_cluster_4_DEA_snRNAseq.xlsx", sheet = "PUT")
te_res_nocov_list_df[["4"]][["PFC"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/HERVs_cluster_4_DEA_snRNAseq.xlsx", sheet = "PFC")
te_res_nocov_list_df[["4"]][["AMY"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/HERVs_cluster_4_DEA_snRNAseq.xlsx", sheet = "AMY")

te_res_nocov_cov_corr_plot_list <- list()
cluster = "4"
for(region in regions){
  signif <- te_res_nocov_list_df [[cluster]][[region]][which(te_res_nocov_list_df[[cluster]][[region]]$padj < 0.05), "TE_id"]
  if(!is.null(te_res_list_df[[cluster]][[region]][te_res_nocov_list_df[[cluster]][[region]]$TE_id,"log2FoldChange"])){
    tmp <- data.frame("TE_id" = te_res_nocov_list_df[[cluster]][[region]]$TE_id,
                    "cov_corr_log2FC" = te_res_list_df[[cluster]][[region]][te_res_nocov_list_df[[cluster]][[region]]$TE_id,"log2FoldChange"],
                    "nocov_log2FC" = te_res_nocov_list_df[[cluster]][[region]]$log2FoldChange,
                    "reported_signif" = te_res_nocov_list_df[[cluster]][[region]]$TE_id %in% signif)
  if(nrow(tmp) > 0){
    te_res_nocov_cov_corr_plot_list[[paste(cluster, region, sep="_")]] <- tmp %>% 
      arrange(reported_signif) %>% 
    ggplot(aes(x = cov_corr_log2FC, y= nocov_log2FC)) + geom_hline(linetype = "dashed", color= "grey", yintercept = 0)  + geom_vline(linetype = "dashed", color= "grey", xintercept = 0)  + geom_point(aes(color=reported_signif, size = reported_signif)) + stat_poly_line(formula = y ~ x) +
    stat_poly_eq(formula = y ~ x, aes(label = after_stat(p.value.label)), label.y = "top") + scale_size_manual(values = c("TRUE" = 1, "FALSE" = 0.3)) + scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "black")) + 
    stat_poly_eq(formula = y ~ x, aes(label = after_stat(adj.rr.label)), label.y = 0.85) +
      theme_bw() + ggtitle(paste(region, cluster)) + labs(x="log2FC (covariate corrected)", y="log2FC (uncorrected)")  
  }  
  }
}

te_res_nocov_cov_corr_plot_list

```

### Now L1s
```{r}
te_res_nocov_list_df <- list()
te_res_nocov_list_df[["4"]] <- list()
te_res_nocov_list_df[["4"]][["SN"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/l1_cluster_4_DEA_snRNAseq.xlsx", sheet = "SN")
te_res_nocov_list_df[["4"]][["PUT"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/l1_cluster_4_DEA_snRNAseq.xlsx", sheet = "PUT")
te_res_nocov_list_df[["4"]][["PFC"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/l1_cluster_4_DEA_snRNAseq.xlsx", sheet = "PFC")
te_res_nocov_list_df[["4"]][["AMY"]] <- openxlsx::read.xlsx("/Volumes/MyPassport/ASAP/code/ASAP_PMDBS_snRNAseq/original/results/tables_vs2/l1_cluster_4_DEA_snRNAseq.xlsx", sheet = "AMY")

te_res_nocov_cov_corr_plot_list <- list()
cluster = "4"
for(region in regions){
  signif <- te_res_nocov_list_df[[cluster]][[region]][which(te_res_nocov_list_df[[cluster]][[region]]$padj < 0.05), "l1_id"]
  if(!is.null(te_res_list_df[[cluster]][[region]][te_res_nocov_list_df[[cluster]][[region]]$l1_id,"log2FoldChange"])){
    tmp <- data.frame("TE_id" = te_res_nocov_list_df[[cluster]][[region]]$l1_id,
                    "cov_corr_log2FC" = te_res_list_df[[cluster]][[region]][te_res_nocov_list_df[[cluster]][[region]]$l1_id,"log2FoldChange"],
                    "nocov_log2FC" = te_res_nocov_list_df[[cluster]][[region]]$log2FoldChange,
                    "reported_signif" = te_res_nocov_list_df[[cluster]][[region]]$l1_id %in% signif)
  if(nrow(tmp) > 0){
    te_res_nocov_cov_corr_plot_list[[paste(cluster, region, sep="_")]] <- tmp %>% 
      arrange(reported_signif) %>% 
    ggplot(aes(x = cov_corr_log2FC, y= nocov_log2FC)) + geom_hline(linetype = "dashed", color= "grey", yintercept = 0)  + geom_vline(linetype = "dashed", color= "grey", xintercept = 0)  + geom_point(aes(color=reported_signif, size = reported_signif)) + stat_poly_line(formula = y ~ x) +
    stat_poly_eq(formula = y ~ x, aes(label = after_stat(p.value.label)), label.y = "top") + scale_size_manual(values = c("TRUE" = 1, "FALSE" = 0.3)) + scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "black")) + 
    stat_poly_eq(formula = y ~ x, aes(label = after_stat(adj.rr.label)), label.y = 0.85) +
      theme_bw() + ggtitle(paste(region, cluster)) + labs(x="log2FC (covariate corrected)", y="log2FC (uncorrected)")  
  }  
  }
}

te_res_nocov_cov_corr_plot_list

```
